---
title: "Pivot Lab"
author: "I 'State Your Name'" #Put your name here, but don't change the other lines in this 'YAML' section of the document
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

You can read more about [R Markdown](http://rmarkdown.rstudio.com) Notebooks. 

Run this chunk to install and/or load rmarkdown (Remember, you can  run this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter* or by pressing the green arrow in the upper right of the chunk.)

```{r}
#install.packages("rmarkdown")
library("rmarkdown")
library("pacman")

pacman::p_load(tidyr, dplyr, janitor, readxl)
```


*After each question, run the chunk to check your code.*

* Don't worry about extracting the data from the PDF. You learned that in Lesson 3. We will begin with the spreadsheet file.

# 1. Use the FraminghamGeneralElectiionResults2017.xlsx spreadsheet file in the data directory to create a new data frame called `myspreadsheet`. 
... Then create a variable called `mycells` that stores the value range of the cells you want to subset from the XLSX file `myspreadsheet` data frame. 
... Then use readxl to create a data frame called `council` that stores just the values of cells A4 to S16 from the XLSX file.
... Then remove the first columns from the council dataframe.
```{r}
myspreadsheet <- 
mycells <- 
council <- 
council <- council[    ]
```

#2. Change the names of the first 18 columns of the `council` data frame so that that the word "Candidate" is pre-pended to all of them.
```{r}

```

#3. When we get to the part in Section 14.7 that begins "To create a 'narrow' tidy data frame with tidyr ..." read this text instead... 


To create a “narrow”“ tidy data frame with tidyr, use the `pivot_wider()` function. I think of it as gathering up all the column information that’s spread out over multiple column names but shouldn’t be, and putting them into a single new column. And in fact, that’s what we want to do here: gather all the precinct information that’s currently scattered across 18 different columns into a single column named “Precinct.”

```{r}
council_long<- council %>% pivot_longer(-Candidate, names_to = "Precinct", values_to = "Votes")
```

You can read more about pivot_wider() and its counterpart pivot_longer() at
* https://www.storybench.org/pivoting-data-from-columns-to-rows-and-back-in-the-tidyverse/
* https://tidyr.tidyverse.org/articles/pivot.html

You will still see tutorials that refrence spread() and gather() and even melt(), but those are all being replaced by uses of pivot_wider and pivot_longer, which professional journalists in the #r channel of newsnerdery.slack.com say is much easier.

# 4. Write code that finds the top finisher in each precinct and stores it in a data frame called `council_winners`.

* Replace the first in Section 14.8 with this: *
The "long" data frame is easier to work with in order to get what we want: showing how each candidate finished in each precinct, and listing the top two candidates per precinct.

```{r}
council_winners <- council_long %>%
  filter(Candidate != "Blanks", Candidate != "Write-Ins") %>% #Frankly, I wouldn't remove these, because it would be news if "Blanks" or "Write-Ins" won the race. We would not want to miss that story.
  group_by(    ) %>%
  
  
  
  
```

# 5. Use the code from #4 to add a column that displays the name of the second-place candidate. (For this one, actually try to get the error message described in the book.)
```{r}








```

# 6. Read http://www.datasciencemadesimple.com/paste-function-in-r/ and describe what the `paste()` function is doing in this code:
```{r}
myresults <- c("Canddiate1", "Candidate 2")
paste(myresults, sep = ",")

#Write your answer here... 





```


# 7. Write code that uses your code from #5 and ensures that in the case of a tie for the top finisher or second-place finisher that the names of all candidates with the same number of votes will be listed. Save the result to a data frame called `council_winners`.
```{r}

```


* In Section 14.9 we are going to use `pivot_wider() `rather than `spread()` * 

So ... 
  tidyr’s pivot_wider() function moves variables from a single column into multiple columns, with each value – in this case candidate – having its own column.

The syntax is pivot_wider(mydf, names_from = "", values_from = ""). So, in this case, we’d want
```{r}
council_table <- council_long %>% pivot_wider(names_from = "Candidate", values_from = "Votes") 

#What we've done here is switched the rows and columns. Before you had candidates as the rows, but now you have precincts as the rows.
```


# 8. Use the `adorn_totals()` function from the `janitor` package to create both column and row totals for the `council_table` data frame. Save these new totals in a new data frame called `council_table_with_totals`
```{r}

#WARNING: I really don't like the idea of doing this unless you are about to publish some data. Creating these column and row totals violates one of the principles of tidy data -- all rows are supposed to be one thing and all the same type of thing, but these sums create a row of aggregate values while all the other rows are precincts.
```


(Don't worry for now about the reference to the DT package. It's a cool tool for creating interactive data tables that you can publish on a web page. But we'll get to that later.)

# 9. Create a final data fram called `by_precinct` in which each row lists the Precinct, the name of the top finisher and the name of the second place finisher (be sure that there is just one row per precinct.)

```{r}

```




On YOUR OWN
#https://s3.amazonaws.com/dl.ncsbe.gov/ENRS/1992_11_03/results_Governor_19921103.pdf

#Run, Knit and Submit
When you've finished, use the "Preview" pulldown menu at the top of this pane to select "Knit to HTML". This will save an HTML file to this project's directory. 

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

After you've created the HTML file of your answers, you can Stage, Commit and then push your .Rmd and .html files back to GitHub where I will grade them.
