---
title: "Pivot Lab"
author: "I 'State Your Name'" #Put your name here, but don't change the other lines in this 'YAML' section of the document
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

You can read more about [R Markdown](http://rmarkdown.rstudio.com) Notebooks. 

Run this chunk to install and/or load rmarkdown (Remember, you can  run this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter* or by pressing the green arrow in the upper right of the chunk.)

```{r}
#install.packages("rmarkdown")
library("rmarkdown")

pacman::p_load(tidyr, dplyr, janitor)
#install.packages("Hmisc")
#unloadNamespace("Hmisc")
```


*After each question, run the chunk to check your code.*



# 1.	Use the as.Date() function from base R to create a variable called `date1` that is a date object representing October 12, 1793
```{r}

```



#Run, Knit and Submit
When you've finished, use the "Preview" pulldown menu at the top of this pane to select "Knit to HTML". This will save an HTML file to this project's directory. 

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

After you've created the HTML file of your answers, you can Stage, Commit and then push your .Rmd and .html files back to GitHub where I will grade them.

